---
title: "Wastewater Surveillance as a Multivariate Change-Point Problem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wastewater Surveillance as a Multivariate Change-Point Problem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 9,
  fig.height = 5,
  message   = FALSE,
  warning   = FALSE
)
```

```{r packages}
library(tidyverse) # for data handling
library(ggplot2) # for plotting
library(knitr) # for rendering vignette + tables
library(canadianmaps) # for mapping canadian cities
library(sf) # for spatial vector data (mapping)
library(maps) # for database of world cities

#multichangepoints from Github
devtools::load_all("~/Dropbox/CANSSI/multi-change-points/")
library(multichangepoints)
set.seed(20260305)
```

## Introduction

Wastewater-based epidemiology (WBE) monitors the concentration of pathogen
genetic material — such as SARS-CoV-2 RNA — in municipal wastewater.
Because viral shedding into sewage can precede clinical case reporting by days
to weeks, surveillance networks that continuously sample and sequence
wastewater at multiple sites can provide an early-warning signal of emerging
or resurging outbreaks, and identify which sites are effected in order to triage resources.

From a statistical perspective, a national WBE network yields a
**multivariate data stream**: at each time point $t$ (here, each
epidemiological week), we observe a measurement $X_{tk}$ at each of
$K$ monitoring sites. The surveillance task has two components:

1. **Change-point detection** — raise an alarm as soon as the data depart
   from a baseline regime, controlling a type-I error criterion such as the
   probability of false alarm (PFA) or the average run length under the null
   (ARL).

2. **Stream identification** — after (or concurrent with) an alarm,
   determine *which* sites $k$ are driving the change, so that public-health
   resources can be directed appropriately.

The `multichangepoints` package is designed for exactly this setting.
A univariate test supermartingale (TSM) is built for each site,
marginal increment sequences are combined across sites
(via `ProductCombiner`, `AverageCombiner`, or `UniversalPortfolioCombiner`),
and a single `ShiryaevRobertsDetector` monitors the combined evidence
sequentially — stopping as soon as the evidence crosses a threshold that
guarantees the desired error control.

---

## Data

The dataset `wastewater_aggregate.csv` is published by the Public Health
Agency of Canada (PHAC), aggregating weekly wastewater measurements from
monitoring sites across Canada.  
Each row contains one observation indexed by `site` by `week` by `pathogen`.  Key columns:

| Column | Description |
|---|---|
| `site` | Monitoring location |
| `province` | Canadian province / territory |
| `weekstart` | Monday of the epidemiological week |
| `measureid` | Pathogen marker (`covN2`, `fluA`, `fluB`, `rsv`) |
| `w_avg` | Weekly average viral signal (copies / mL, log10-transformed at source) |

```{r load-data}
#NB: data dictionary says empty sites indicate aggregates, but there are no empty sites
ww <- read_csv(
  here::here("Data", "wastewater_aggregate.csv"),
  show_col_types = FALSE
) %>%
  mutate(weekstart = as.Date(weekstart))
```

```{r head}
head(ww)
```

The data span `r format(min(ww$weekstart), "%B %Y")` to
`r format(max(ww$weekstart), "%B %Y")`, cover
`r length(unique(ww$site[ww$site != ""]))` monitoring sites across
`r length(unique(ww$province[ww$province != ""]))` provinces and territories,
and include four pathogen markers:
`r paste(unique(ww$measureid), collapse = ", ")`.

There are `length(unique(ww$site))` sites in total, with varying numbers of observations depending on when monitoring started and whether they are missing any data.


---

## COVID-19 (covN2)

We filter to the SARS-CoV-2 N2 gene target and remove the national-aggregate
rows (blank site name), retaining only site-level observations.

```{r filter-covid}
cov <- ww %>%
  filter(measureid == "covN2", site != "") %>%
  mutate(province = if_else(province == "", NA_character_, province))

cat(
  "Sites:", n_distinct(cov$site), "\n",
  "Provinces:", n_distinct(cov$province, na.rm = TRUE), "\n",
  "Weeks:", n_distinct(cov$weekstart), "\n",
  "Rows:", nrow(cov), "\n"
)
```


### Sampling locations and number of observations

Sites vary in their number of observations primarily because monitoring started at different times at different sites, but also because some weeks are missing measurements.

```{r observations per site, fig.height = 6, fig.width = 9, fig.cap = "Canadian wastewater monitoring sites. Point size reflects the number of observations of COVID-19 viral loads; province abbreviations from canadianmaps."}

# ---- observation counts per site -------------------------------------------
site_counts <- ww %>%
  filter(site != "", measureid == "covN2") %>%
  group_by(site, city, province) %>%
  summarize(count = n()) %>%
  mutate(city_clean = trimws(city))

# ---- geocode: auto-match city names from maps::world.cities ----------------
ca_cities <- world.cities[world.cities$country.etc == "Canada", ]

# joins cities by name and take largest if there are any repeats
site_geo <- site_counts %>%
  left_join(
    ca_cities %>% select(name, lat, long, pop),
    by = c("city_clean" = "name"),
    relationship = "many-to-many"
  ) %>%
  group_by(site) %>%
  slice_max(pop, n = 1, with_ties = FALSE) %>%
  ungroup()

# ---- manual coordinates for the 18 unmatched sites -------------------------
manual_coords <- tribble(
  ~site,                                        ~lat,    ~long,
  "Battery Point",                              46.14,  -60.19,
  "Battleford",                                 52.73, -108.30,
  "Birch Hills",                                52.98, -105.44,
  "City of Charlottetown and Town of Stratford",46.24,  -63.13,
  "Dominion-Bridgeport",                        46.13,  -60.00,
  "Miramichi",                                  47.02,  -65.50,
  "Pasqua",                                     50.58, -103.98,
  "Peel G.E. Booth",                            43.73,  -79.72,
  "Southey",                                    50.95, -104.49,
  "St. John's",                                 47.56,  -52.71,
  "St. Stephen",                                45.20,  -67.28,
  "Trenton",                                    45.61,  -62.63,
  "Vancouver Annacis Island",                   49.17, -122.92,
  "Vancouver Iona Island",                      49.22, -123.21,
  "Vancouver Lions Gate",                       49.35, -123.13,
  "Vancouver Lulu Island",                      49.14, -123.19,
  "Vancouver Northwest Langley",                49.17, -122.60,
  "\u00cele-\u00e0-la-Crosse",                  55.45, -107.89
)

site_geo <- site_geo %>%
  rows_update(manual_coords, by = "site", unmatched = "ignore") %>%
  filter(!is.na(lat))

# ---- project points into the canadianmaps Lambert CRS ----------------------
lcc_crs <- "+proj=lcc +lat_1=49 +lat_2=77 +lon_0=-91.52 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"

sites_sf <- site_geo %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326) %>%
  st_transform(crs = lcc_crs)

# ---- map -------------------------------------------------------------------
ggplot() +
  geom_prov(fill = "gray93", colour = "gray60", size = 0.25) +
  geom_sf(
    data   = sites_sf,
    aes(size = count),
    colour = "#1A5276",
    alpha  = 0.75,
    shape  = 16
  ) +
  text_prov(colour = "gray40", size = 2.8) +
  scale_size_area(
    name     = "Weeks\nobserved",
    max_size = 8,
    breaks   = c(50, 100, 150, 200)
  ) +
  crs_coord() +
  labs(
    title   = "Canadian wastewater monitoring network for COVID-19",
    caption = "Point size = number of weekly COVID-19 observations per site."
  ) +
  theme_map() +
  theme(
    legend.position  = c(0.93, 0.18),
    legend.background = element_rect(fill = alpha("white", 0.7), colour = NA),
    legend.title     = element_text(size = 9),
    plot.title       = element_text(face = "bold", margin = margin(b = 4)),
    plot.caption     = element_text(colour = "gray50", size = 8,
                                    margin = margin(t = 4))
  )
```


### Trend overview — all sites

The plot below shows weekly COVID signal (`w_avg`) at every monitoring site
as a thin semi-transparent line, with the national average overlaid in
bold.  Vertical dashed lines mark approximate Omicron wave peaks for
reference.

```{r plot-all-sites, fig.height = 4.5, fig.cap = "Weekly SARS-CoV-2 wastewater signal at all Canadian monitoring sites (COVID). Each gray line is one site; the national aggregate (bold blue) reflects the country-wide trajectory."}

national <- ww %>%
  filter(measureid == "covN2") %>%
  group_by(weekstart) %>%
  summarize(w_avg = mean(w_avg)) %>%
  mutate(province = "National")

ggplot() +
  # Individual sites
  geom_line(
    data  = cov,
    aes(x = weekstart, y = w_avg, group = site),
    colour = "gray60", alpha = 0.35, linewidth = 0.35
  ) +
  # National aggregate
  geom_line(
    data      = national,
    aes(x = weekstart, y = w_avg),
    colour    = "#1A5276",
    linewidth = 1.1
  ) +
  # Approximate Omicron wave peaks (BA.1, BA.2, BA.4/5)
  geom_vline(
    xintercept = as.Date(c("2022-01-09", "2022-04-03", "2022-07-17")),
    linetype   = "dashed",
    colour     = "firebrick",
    alpha      = 0.55
  ) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x       = NULL,
    y       = "Weekly average signal (copies / mL)",
    title   = "SARS-CoV-2 wastewater signal — all Canadian sites",
    caption = "Gray: individual sites.  Bold blue: national aggregate.
Dashed red: approximate Omicron BA.1 / BA.2 / BA.4-5 wave peaks (source: https://onlinelibrary.wiley.com/doi/10.1111/irv.70197)."
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 30, hjust = 1),
    plot.caption    = element_text(colour = "gray40", size = 8),
    panel.grid.minor = element_blank()
  )
```

Covid was detected in `r round(100 * mean(cov$w_avg > 0), 1)`% of weeks.

### Trends by province

With 76 sites on a single panel the regional variation is hard to see.
The faceted plot below shows one line per site within each province,
making it easier to assess synchrony and heterogeneity of outbreaks.

```{r plot-by-province, fig.height = 9, fig.width = 10, fig.cap = "SARS-CoV-2 wastewater signal by province. Each line is one monitoring site; colour distinguishes sites within a province."}

# Provinces with at least some data — ordered W → E then territories
province_order <- c(
  "British Columbia", "Alberta", "Saskatchewan", "Manitoba",
  "Ontario", "Quebec",
  "New Brunswick", "Nova Scotia", "Prince Edward Island",
  "Newfoundland and Labrador",
  "Yukon"
)

cov_prov <- cov %>%
  filter(!is.na(province)) %>%
  mutate(province = factor(province, levels = province_order))

ggplot(cov_prov, aes(x = weekstart, y = w_avg,
                     colour = site, group = site)) +
  geom_line(linewidth = 0.55, alpha = 0.8) +
  facet_wrap(~ province, ncol = 3, scales = "free_y") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  labs(
    x     = NULL,
    y     = "Weekly average signal (copies / mL)",
    title = "SARS-CoV-2 wastewater signal by province",
    subtitle = "Each line = one monitoring site  |  y-axis free within panel"
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 8),
    strip.background = element_rect(fill = "#D6EAF8"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )
```

The wave structure is broadly synchronous across provinces — consistent with
a national-level change in pathogen prevalence — but the magnitude and
timing of individual site peaks vary, reflecting local population sizes,
sampling frequency, and catchment characteristics.  This heterogeneity
motivates a **multi-stream approach**: rather than pooling sites into a
single signal, the `multichangepoints` framework tests each site
sequentially and combines the evidence in a principled way, allowing the
detector to remain powerful even when only a subset of sites shows an
elevated signal.


### Missing Data

Some sites are missing weeks, wherein COVID-19 measurements were not taken.

```{r missing data}
missing_by_site <- cov %>%
  group_by(site) %>%
  summarize(
    first_week = min(weekstart),
    last_week  = max(weekstart),
    observed   = list(sort(unique(weekstart))),
    .groups    = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    expected           = list(seq(first_week, last_week, by = "7 days")),
    n_expected_observed = length(expected),
    n_missing          = sum(!expected %in% observed),
    pct_missing        = round(100 * n_missing / n_expected_observed, 1)
  ) %>%
  ungroup() %>%
  select(site, n_expected_observed, n_missing, pct_missing) %>%
  arrange(desc(pct_missing))

head(missing_by_site, 10)

summary(missing_by_site$pct_missing)
```


# Change-point detection for WBE

Change-point detection with WBE data is a major challenge because it is very difficult to identify the _state of control_ that is supposed to be represented by the pre-change model. 
In this case, the state of control is a site-specific simple or composite model from which the observed sequence is generated when there is no epidemic.
Since we have multiple sites, we might try to make some kind of 'synthetic state of control' by synthesizing data across sites that were epidemic-free at various times. 
Unfortunately, this is subject to the many uncertainties inherent in wastewater surveillance across multiple sites [Joung et al (2023)](https://pmc.ncbi.nlm.nih.gov/articles/PMC10890812/pdf/CCDR-49-166.pdf), including:

- **Target quantity**: The true target quantity of interest is the average viral load in a particular population week over week. This quantity varies naturally over time when the virus is in an endemic state. While this variation is not genuinely random, it can be modeled as such. For instance, we might assume it maintains some low-level average with as-if-Gaussian fluctuations from week to week, in which case the sequence is _stationary_ in the endemic phase.
- **Sampling**: The methods of sample collection, shipping, and storage (compositing vs grab sampling, location, timing, shipping temperature, distance to lab, missing data), which may vary across sites and time.
- **Analysis**: Laboratory analysis variability (de-gritting subsampling, homogenization, instrument error, human error).
- **Infrastructure**: The location and engineering of sewage systems and wastewater treatment plants (catchment area size and population, combined or separate ('sanitary') collection of runoff and wastewater).
- **Weather**: Weather conditions (rainwater or snowmelt contributing to influent, temperature, pH), which vary across sites and seasons, with some degree of stochasticity. 

In the literature, these uncertainties are dealt with under the rubric of _normalization_, which is an active and un-clarified area of research. 
In some sense, defining a proper state of control from which deviations are meaningful is the whole challenge of WBE. 
It may be easier to define the state of control for certain sites, like those with sanitary sewage systems that are less influenced by major weather events. 


For any given site, we could imagine an ideal way to establish its state of control via _potential outcomes_. In one potential outcome, the site experiences a COVID epidemic and in another epidemic the site experiences no COVID epidemic (or some kind of endemic state, both of these are very rough definitions). 
This is, of course, impossible in reality; we cannot even imagine a reasonable experiment because there is no way to manipulate the hypothetical intervention 'COVID epidemic'. 
This could be considered a problem in _observational causal inference_. 
One way to establish the state of control is through some kind of matching, wherein we find pairs site that are similar along various axes likely to effect background rates (collection, engineering, weather, etc); but in one site there is no epidemic and in another site there is an epidemic. 
We could also do this by site-periods, with sites serving as their own control over periods where they are epidemic free.
Another way is to fit models for the state of control. 
This is attempted below. 
In either case, establishing a meaningful state of control (a simple or composite pre-change model) should be done separately from running a change-point algorithm, which can detect deviations but cannot define the state of control without a user. 


## Example model and CPD

This section is a placeholder.


```{r remove trend and model as AR(1)}
# 1. start with a single site with no missing data (Vancouver)
site_series <- cov %>%
  filter(site == "Vancouver Lions Gate") %>%
  arrange(weekstart) %>%
  mutate(y = log1p(w_avg)) %>%
  mutate(log_change = y - lag(y)) %>%
  filter(!is.na(log_change))

x <- site_series$log_change # time series is the log change in concentrations

# 2. Fit AR(2) using arima() (degree determined by AIC)
fit <- arima(x, order = c(2, 0, 0))
resid <- residuals(fit)

# Fit diagnostics
Box.test(resid, lag = 10, type = "Ljung-Box")
par(mfrow = c(2, 2))
plot(resid, type = "l", main = "Residuals over time")
acf(resid, main = "ACF of residuals")
pacf(resid, main = "PACF of residuals")
qqnorm(resid); qqline(resid)
```

```{r detector on residuals}
# point estimate of nuisance parameter 
sigma_hat <- fit$sigma2
model <- GaussianModel(
  mean_pre = 0,
  sd_pre = sigma_hat,
  mean_post = c(0, 1), # this captures a one-sided alternative, that the rate of change increases
  sd_post = sigma_hat,
  method = "predictable"
)
inc <- compute_increments(TSM(model), x, log = TRUE)
sr <- ShiryaevRobertsDetector(
  alpha = 1/52, # average run length without false alarm is 1 year
  criterion = "ARL"
)
det_out <- run_detector(sr, inc, log = TRUE)
```

```{r plot detector}
par(mfrow = c(2, 2))
plot(site_series$w_avg, x = site_series$weekstart, type = 'l', lwd = 2, col = 'darkorange3', main = "Original Vancouver COVID WBE Data")
plot(x, type = 'l', x = site_series$weekstart, lwd = 2, col = 'forestgreen', main = "Transformed Data (log-change)")


z <- increments_to_tsm(inc, log = TRUE)
plot(z, type = "l", lwd = 2, col = "steelblue",
     main = "log-TSM sequence", xlab = "Week", x = site_series$weekstart, ylab = "TSM")

plot(det_out$statistic, type = "l", lwd = 2, col = "firebrick",
     main = "log-S-R statistic (ARL = 20)", xlab = "Week", x = site_series$weekstart, ylab = "R_t")
abline(h = log(sr@threshold), lty = 3, col = "gray50")
if (det_out$alarm) {
  abline(v = site_series$weekstart[det_out$stopping_time], lwd = 2, col = "darkgreen")
}
par(mfrow = c(1,1))
```

## Joint wastewater site monitoring

To pass the wastewater data through the CPD algorithm, we need the site data as a matrix with `NA` where there are missing values. 

```{r define joint matrix}
# Full weekly spine: every 7 days from global min to global max
all_weeks <- seq(min(cov$weekstart), max(cov$weekstart), by = "7 days")

joint_matrix <- cov %>%
  mutate(log_change = log1p(w_avg) - log1p(lag(w_avg))) %>%
  select(site, weekstart, log_change) %>%
  complete(site, weekstart = all_weeks) %>%   # pad all site × week combinations with NA
  arrange(site, weekstart) %>%
  pivot_wider(names_from = weekstart, values_from = log_change) %>%
  column_to_rownames("site") %>%
  as.matrix()

cat("Dimensions: ", nrow(joint_matrix), "sites x", ncol(joint_matrix), "weeks\n")
cat("Proportion NA: ", round(mean(is.na(joint_matrix)), 3), "\n")
# white spaces are missing values
image(t(joint_matrix), ylab = "Sites", xlab = "Weeks", xaxt = "n", yaxt = "n") 
```


Next steps: 

- Handle gaps (missing data)
- Allow repeat alarms (via resets)
- Multiple testing error control (FWER, FDR)
- Define appropriate state of control (pre-change model) on holdout. This is basically a causal inference problem: imagine a _potential sequence_ with no epidemic and a potential sequence with epidemic for each site. We observe an _interleaving_  (almost like a cross-over design) of these sequences, with measurement error, and wish to identify when we switch between these states with little delay and with continuous monitoring. 

