---
title: "Wastewater Surveillance as a Multivariate Change-Point Problem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wastewater Surveillance as a Multivariate Change-Point Problem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 9,
  fig.height = 5,
  message   = FALSE,
  warning   = FALSE
)
```

```{r packages}
library(tidyverse)
library(ggplot2)
library(knitr)
```

## Introduction

Wastewater-based epidemiology (WBE) monitors the concentration of pathogen
genetic material — such as SARS-CoV-2 RNA — in municipal wastewater.
Because viral shedding into sewage can precede clinical case reporting by days
to weeks, surveillance networks that continuously sample and sequence
wastewater at multiple sites can provide an early-warning signal of emerging
or resurging outbreaks, and identify which sites are effected in order to triage resources.

From a statistical perspective, a national WBE network yields a
**multivariate data stream**: at each time point $t$ (here, each
epidemiological week), we observe a measurement $X_{tk}$ at each of
$K$ monitoring sites. The surveillance task has two components:

1. **Change-point detection** — raise an alarm as soon as the data depart
   from a baseline regime, controlling a type-I error criterion such as the
   probability of false alarm (PFA) or the average run length under the null
   (ARL).

2. **Stream identification** — after (or concurrent with) an alarm,
   determine *which* sites $k$ are driving the change, so that public-health
   resources can be directed appropriately.

The `multichangepoints` package is designed for exactly this setting.
A univariate test supermartingale (TSM) is built for each site,
marginal increment sequences are combined across sites
(via `ProductCombiner`, `AverageCombiner`, or `UniversalPortfolioCombiner`),
and a single `ShiryaevRobertsDetector` monitors the combined evidence
sequentially — stopping as soon as the evidence crosses a threshold that
guarantees the desired error control.

---

## Data

The dataset `wastewater_aggregate.csv` is published by the Public Health
Agency of Canada, and aggregating weekly wastewater measurements from
monitoring sites across Canada.  
Each row contains one observation indexed by `site` $\times$ `week` $\times$ `pathogen`.  Key columns:

| Column | Description |
|---|---|
| `site` | Monitoring location |
| `province` | Canadian province / territory |
| `weekstart` | Monday of the epidemiological week |
| `measureid` | Pathogen marker (`covN2`, `fluA`, `fluB`, `rsv`) |
| `w_avg` | Weekly average viral signal (copies / mL, log10-transformed at source) |

```{r load-data}
#NB: data dictionary says empty sites indicate aggregates, but there are no empty sites
ww <- read_csv(
  here::here("Data", "wastewater_aggregate.csv"),
  show_col_types = FALSE
) %>%
  mutate(weekstart = as.Date(weekstart))
```

```{r head}
head(ww)
```

The data span `r format(min(ww$weekstart), "%B %Y")` to
`r format(max(ww$weekstart), "%B %Y")`, cover
`r length(unique(ww$site[ww$site != ""]))` monitoring sites across
`r length(unique(ww$province[ww$province != ""]))` provinces and territories,
and include four pathogen markers:
`r paste(unique(ww$measureid), collapse = ", ")`.

There are `length(unique(ww$site))` sites in total, with varying numbers of observations depending on when monitoring started and whether they are missing any data.


---

## COVID-19 (covN2)

We filter to the SARS-CoV-2 N2 gene target and remove the national-aggregate
rows (blank site name), retaining only site-level observations.

```{r filter-covid}
cov <- ww %>%
  filter(measureid == "covN2", site != "") %>%
  mutate(province = if_else(province == "", NA_character_, province))

cat(
  "Sites:", n_distinct(cov$site), "\n",
  "Provinces:", n_distinct(cov$province, na.rm = TRUE), "\n",
  "Weeks:", n_distinct(cov$weekstart), "\n",
  "Rows:", nrow(cov), "\n"
)
```

### Number of observations

Sites vary in their number of observations primarily because monitoring started at different times at different sites, but also because some weeks are missing measurements.

```{r observations per site, fig.height = 3.5, fig.cap = "Distribution of observation counts per monitoring site across all pathogens."}
site_counts <- ww %>%
  filter(site != "") %>%
  filter(measureid == "covN2") %>%
  group_by(site) %>%
  summarize(count = n(), .groups = "drop")

ggplot(site_counts, aes(x = count)) +
  geom_histogram(
    binwidth = 25,
    fill     = "#2E86C1",
    colour   = "white",
    linewidth = 0.3
  ) +
  geom_vline(
    xintercept = median(site_counts$count),
    linetype   = "dashed",
    colour     = "firebrick",
    linewidth  = 0.8
  ) +
  annotate(
    "text",
    x      = median(site_counts$count) + 8,
    y      = Inf,
    label  = paste0("median = ", median(site_counts$count)),
    vjust  = 1.5,
    hjust  = 0,
    colour = "firebrick",
    size   = 3.5
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.02))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    x     = "Weeks observed per site (all pathogens)",
    y     = "Number of sites",
    title = "How many observations does each site have?"
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```


### Trend overview — all sites

The plot below shows weekly COVID signal (`w_avg`) at every monitoring site
as a thin semi-transparent line, with the national average overlaid in
bold.  Vertical dashed lines mark approximate Omicron wave peaks for
reference.

```{r plot-all-sites, fig.height = 4.5, fig.cap = "Weekly SARS-CoV-2 wastewater signal at all Canadian monitoring sites (covN2). Each gray line is one site; the national aggregate (bold blue) reflects the country-wide trajectory."}

national <- ww %>%
  group_by(weekstart) %>%
  summarize(w_avg = mean(w_avg)) %>%
  mutate(province = "National")

ggplot() +
  # Individual sites
  geom_line(
    data  = cov,
    aes(x = weekstart, y = w_avg, group = site),
    colour = "gray60", alpha = 0.35, linewidth = 0.35
  ) +
  # National aggregate
  geom_line(
    data      = national,
    aes(x = weekstart, y = w_avg),
    colour    = "#1A5276",
    linewidth = 1.1
  ) +
  # Approximate Omicron wave peaks (BA.1, BA.2, BA.4/5)
  geom_vline(
    xintercept = as.Date(c("2022-01-09", "2022-04-03", "2022-07-17")),
    linetype   = "dashed",
    colour     = "firebrick",
    alpha      = 0.55
  ) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x       = NULL,
    y       = "Weekly average signal (copies / mL)",
    title   = "SARS-CoV-2 wastewater signal — all Canadian sites",
    caption = "Gray: individual sites.  Bold blue: national aggregate.
Dashed red: approximate Omicron BA.1 / BA.2 / BA.4-5 wave peaks (source: https://onlinelibrary.wiley.com/doi/10.1111/irv.70197)."
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 30, hjust = 1),
    plot.caption    = element_text(colour = "gray40", size = 8),
    panel.grid.minor = element_blank()
  )
```

### Trends by province

With 76 sites on a single panel the regional variation is hard to see.
The faceted plot below shows one line per site within each province,
making it easier to assess synchrony and heterogeneity of outbreaks.

```{r plot-by-province, fig.height = 9, fig.width = 10, fig.cap = "SARS-CoV-2 wastewater signal by province. Each line is one monitoring site; colour distinguishes sites within a province."}

# Provinces with at least some data — ordered W → E then territories
province_order <- c(
  "British Columbia", "Alberta", "Saskatchewan", "Manitoba",
  "Ontario", "Quebec",
  "New Brunswick", "Nova Scotia", "Prince Edward Island",
  "Newfoundland and Labrador",
  "Yukon"
)

cov_prov <- cov %>%
  filter(!is.na(province)) %>%
  mutate(province = factor(province, levels = province_order))

ggplot(cov_prov, aes(x = weekstart, y = w_avg,
                     colour = site, group = site)) +
  geom_line(linewidth = 0.55, alpha = 0.8) +
  facet_wrap(~ province, ncol = 3, scales = "free_y") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  labs(
    x     = NULL,
    y     = "Weekly average signal (copies / mL)",
    title = "SARS-CoV-2 wastewater signal by province",
    subtitle = "Each line = one monitoring site  |  y-axis free within panel"
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 8),
    strip.background = element_rect(fill = "#D6EAF8"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )
```

The wave structure is broadly synchronous across provinces — consistent with
a national-level change in pathogen prevalence — but the magnitude and
timing of individual site peaks vary, reflecting local population sizes,
sampling frequency, and catchment characteristics.  This heterogeneity
motivates a **multi-stream approach**: rather than pooling sites into a
single signal, the `multichangepoints` framework tests each site
sequentially and combines the evidence in a principled way, allowing the
detector to remain powerful even when only a subset of sites shows an
elevated signal.
