---
title: "Wastewater Surveillance as a Multivariate Change-Point Problem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wastewater Surveillance as a Multivariate Change-Point Problem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 9,
  fig.height = 5,
  message   = FALSE,
  warning   = FALSE
)
```

```{r packages}
library(tidyverse) # for data handling
library(ggplot2) # for plotting
library(knitr) # for rendering vignette + tables
library(canadianmaps) # for mapping canadian cities
library(sf) # for spatial vector data (mapping)
library(maps) # for database of world cities
library(naniar) # for missing data visualization

#multichangepoints from Github
devtools::load_all("~/Dropbox/CANSSI/multi-change-points/")
library(multichangepoints)
set.seed(20260305)
```

## Introduction

Wastewater-based epidemiology (WBE) monitors the concentration of pathogen
genetic material — such as SARS-CoV-2 RNA — in municipal wastewater.
Because viral shedding into sewage can precede clinical case reporting by days
to weeks [Peng et al (2025)](https://www.sciencedirect.com/science/article/pii/S1755436525000283), surveillance networks that continuously sample and sequence wastewater at multiple sites can provide early-warning signals of emerging or resurging outbreaks, and identify which sites are effected in order to triage resources.

From a statistical perspective, a national WBE network yields a
**multivariate data stream**: at each time point $t$ (here, each
epidemiological week), we observe a measurement $X_{tk}$ at each of
$K$ monitoring sites. The surveillance task has two components:

1. **Change-point detection** — raise an alarm as soon as the data depart
   from a baseline regime, controlling a type-I error criterion such as the
   probability of false alarm (PFA) or the average run length under the null
   (ARL).

2. **Stream identification** — after (or concurrent with) an alarm,
   determine *which* sites $k$ are driving the change, so that public-health
   resources can be directed appropriately.

The `multichangepoints` package is designed for exactly this setting.
A univariate test supermartingale (TSM) is built for each site,
marginal increment sequences are combined across sites
(via `ProductCombiner`, `AverageCombiner`, or `UniversalPortfolioCombiner`),
and a single `ShiryaevRobertsDetector` monitors the combined evidence
sequentially — stopping as soon as the evidence crosses a threshold that
guarantees the desired error control.

---

## Data

The dataset `wastewater_aggregate.csv` is published by the Public Health
Agency of Canada (PHAC), aggregating weekly wastewater measurements from
monitoring sites across Canada.  
Each row contains one observation indexed by `site` by `week` by `pathogen`.  Key columns:

| Column | Description |
|---|---|
| `site` | Monitoring location |
| `province` | Canadian province / territory |
| `weekstart` | Monday of the epidemiological week |
| `measureid` | Pathogen marker (`covN2`, `fluA`, `fluB`, `rsv`) |
| `covN2` | Weekly average viral signal (copies / mL, log10-transformed at source) |

```{r load-data}
#NB: data dictionary says empty sites indicate aggregates, but there are no empty sites
ww <- read_csv(
  here::here("Data", "wastewater_aggregate.csv"),
  show_col_types = FALSE
) %>%
  mutate(weekstart = as.Date(weekstart))
```

```{r head}
head(ww)
```

The data span `r format(min(ww$weekstart), "%B %Y")` to
`r format(max(ww$weekstart), "%B %Y")`, cover
`r length(unique(ww$site[ww$site != ""]))` monitoring sites across
`r length(unique(ww$province[ww$province != ""]))` provinces and territories,
and include four pathogen markers:
`r paste(unique(ww$measureid), collapse = ", ")`.

There are `length(unique(ww$site))` sites in total, with varying numbers of observations depending on when monitoring started and whether they are missing any data.


---

## COVID-19 (covN2)

We filter to the SARS-CoV-2 N2 gene target and remove the national-aggregate
rows (blank site name), retaining only site-level observations.

```{r filter to covid but include ratio with rsv}
cov <- ww %>%
  select(Location, site, province, country, weekstart, measureid, w_avg) %>%
  filter(measureid %in% c("covN2", "rsv"), site != "") %>%
  mutate(province = if_else(province == "", NA_character_, province)) %>%
  pivot_wider(values_from = "w_avg", names_from = "measureid") %>%
  mutate(covid_rsv_ratio = covN2/rsv)

cat(
  "Sites:", n_distinct(cov$site), "\n",
  "Provinces:", n_distinct(cov$province, na.rm = TRUE), "\n",
  "Weeks:", n_distinct(cov$weekstart), "\n",
  "Rows:", nrow(cov), "\n"
)
```


### Sampling locations and number of observations

Sites vary in their number of observations primarily because monitoring started at different times at different sites, but also because some weeks are missing measurements.

```{r observations per site, fig.height = 6, fig.width = 9, fig.cap = "Canadian wastewater monitoring sites. Point size reflects the number of observations of COVID-19 viral loads; province abbreviations from canadianmaps."}

# ---- observation counts per site -------------------------------------------
site_counts <- ww %>%
  filter(site != "", measureid %in% c("covN2")) %>%
  group_by(site, city, province) %>%
  summarize(count = n()) %>%
  mutate(city_clean = trimws(city)) 

# ---- geocode: auto-match city names from maps::world.cities ----------------
ca_cities <- world.cities[world.cities$country.etc == "Canada", ]

# joins cities by name and take largest if there are any repeats
site_geo <- site_counts %>%
  left_join(
    ca_cities %>% select(name, lat, long, pop),
    by = c("city_clean" = "name"),
    relationship = "many-to-many"
  ) %>%
  group_by(site) %>%
  slice_max(pop, n = 1, with_ties = FALSE) %>%
  ungroup()

# ---- manual coordinates for the 18 unmatched sites -------------------------
manual_coords <- tribble(
  ~site,                                        ~lat,    ~long,
  "Battery Point",                              46.14,  -60.19,
  "Battleford",                                 52.73, -108.30,
  "Birch Hills",                                52.98, -105.44,
  "City of Charlottetown and Town of Stratford",46.24,  -63.13,
  "Dominion-Bridgeport",                        46.13,  -60.00,
  "Miramichi",                                  47.02,  -65.50,
  "Pasqua",                                     50.58, -103.98,
  "Peel G.E. Booth",                            43.73,  -79.72,
  "Southey",                                    50.95, -104.49,
  "St. John's",                                 47.56,  -52.71,
  "St. Stephen",                                45.20,  -67.28,
  "Trenton",                                    45.61,  -62.63,
  "Vancouver Annacis Island",                   49.17, -122.92,
  "Vancouver Iona Island",                      49.22, -123.21,
  "Vancouver Lions Gate",                       49.35, -123.13,
  "Vancouver Lulu Island",                      49.14, -123.19,
  "Vancouver Northwest Langley",                49.17, -122.60,
  "\u00cele-\u00e0-la-Crosse",                  55.45, -107.89
)

site_geo <- site_geo %>%
  rows_update(manual_coords, by = "site", unmatched = "ignore") %>%
  filter(!is.na(lat))

# ---- project points into the canadianmaps Lambert CRS ----------------------
lcc_crs <- "+proj=lcc +lat_1=49 +lat_2=77 +lon_0=-91.52 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"

sites_sf <- site_geo %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326) %>%
  st_transform(crs = lcc_crs)

# ---- map -------------------------------------------------------------------
ggplot() +
  geom_prov(fill = "gray93", colour = "gray60", size = 0.25) +
  geom_sf(
    data   = sites_sf,
    aes(size = count),
    colour = "#1A5276",
    alpha  = 0.75,
    shape  = 16
  ) +
  text_prov(colour = "gray40", size = 2.8) +
  scale_size_area(
    name     = "Weeks\nobserved",
    max_size = 8,
    breaks   = c(50, 100, 150, 200)
  ) +
  crs_coord() +
  labs(
    title   = "Canadian wastewater monitoring network for COVID-19",
    caption = "Point size = number of weekly COVID-19 observations per site."
  ) +
  theme_map() +
  theme(
    legend.position  = c(0.93, 0.18),
    legend.background = element_rect(fill = alpha("white", 0.7), colour = NA),
    legend.title     = element_text(size = 9),
    plot.title       = element_text(face = "bold", margin = margin(b = 4)),
    plot.caption     = element_text(colour = "gray50", size = 8,
                                    margin = margin(t = 4))
  )
```


### Trend overview — all sites

The plot below shows weekly COVID signal (`covN2`) at every monitoring site
as a thin semi-transparent line, with the national average overlaid in
bold.  Vertical dashed lines mark approximate Omicron wave peaks for
reference.

```{r plot-all-sites, fig.height = 4.5, fig.cap = "Weekly SARS-CoV-2 wastewater signal at all Canadian monitoring sites (COVID). Each gray line is one site; the national aggregate (bold blue) reflects the country-wide trajectory."}

national <- ww %>%
  filter(measureid == "covN2") %>%
  group_by(weekstart) %>%
  summarize(covN2 = mean(w_avg)) %>%
  mutate(province = "National")

ggplot() +
  # Individual sites
  geom_line(
    data  = cov,
    aes(x = weekstart, y = covN2, group = site),
    colour = "gray60", alpha = 0.35, linewidth = 0.35
  ) +
  # National aggregate
  geom_line(
    data      = national,
    aes(x = weekstart, y = covN2),
    colour    = "#1A5276",
    linewidth = 1.1
  ) +
  # Approximate Omicron wave peaks (BA.1, BA.2, BA.4/5)
  geom_vline(
    xintercept = as.Date(c("2022-01-09", "2022-04-03", "2022-07-17")),
    linetype   = "dashed",
    colour     = "firebrick",
    alpha      = 0.55
  ) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x       = NULL,
    y       = "Weekly average signal (copies / mL)",
    title   = "SARS-CoV-2 wastewater signal — all Canadian sites",
    caption = "Gray: individual sites.  Bold blue: national aggregate.
Dashed red: approximate Omicron BA.1 / BA.2 / BA.4-5 wave peaks (source: https://onlinelibrary.wiley.com/doi/10.1111/irv.70197)."
  ) +
  theme_bw(base_size = 11) +
  theme(
    axis.text.x     = element_text(angle = 30, hjust = 1),
    plot.caption    = element_text(colour = "gray40", size = 8),
    panel.grid.minor = element_blank()
  )
```

Covid was detected in `r round(100 * mean(cov$covN2 > 0), 1)`% of weeks.

### Trends by province

With 76 sites on a single panel the regional variation is hard to see.
The faceted plot below shows one line per site within each province,
making it easier to assess synchrony and heterogeneity of outbreaks.

```{r plot-by-province, fig.height = 9, fig.width = 10, fig.cap = "SARS-CoV-2 wastewater signal by province. Each line is one monitoring site; colour distinguishes sites within a province."}

# Provinces with at least some data — ordered W → E then territories
province_order <- c(
  "British Columbia", "Alberta", "Saskatchewan", "Manitoba",
  "Ontario", "Quebec",
  "New Brunswick", "Nova Scotia", "Prince Edward Island",
  "Newfoundland and Labrador",
  "Yukon"
)

cov_prov <- cov %>%
  filter(!is.na(province)) %>%
  mutate(province = factor(province, levels = province_order))

ggplot(cov_prov, aes(x = weekstart, y = covN2,
                     colour = site, group = site)) +
  geom_line(linewidth = 0.55, alpha = 0.8) +
  facet_wrap(~ province, ncol = 3, scales = "free_y") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  labs(
    x     = NULL,
    y     = "Weekly average signal (copies / mL)",
    title = "SARS-CoV-2 wastewater signal by province",
    subtitle = "Each line = one monitoring site  |  y-axis free within panel"
  ) +
  theme_bw(base_size = 10) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1, size = 8),
    strip.background = element_rect(fill = "#D6EAF8"),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank()
  )
```

The wave structure is broadly synchronous across provinces — consistent with
a national-level change in pathogen prevalence — but the magnitude and
timing of individual site peaks vary, reflecting local population sizes,
sampling frequency, and catchment characteristics.  This heterogeneity
motivates a **multi-stream approach**: rather than pooling sites into a
single signal, the `multichangepoints` framework tests each site
sequentially and combines the evidence in a principled way, allowing the
detector to remain powerful even when only a subset of sites shows an
elevated signal.


### Missing Data

Some sites are missing weeks, wherein COVID-19 measurements were not taken.

```{r missing data}
missing_by_site <- cov %>%
  group_by(site) %>%
  summarize(
    first_week = min(weekstart),
    last_week  = max(weekstart),
    observed   = list(sort(unique(weekstart))),
    .groups    = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    expected           = list(seq(first_week, last_week, by = "7 days")),
    n_expected_observed = length(expected),
    n_missing          = sum(!expected %in% observed),
    pct_missing        = round(100 * n_missing / n_expected_observed, 1)
  ) %>%
  ungroup() %>%
  select(site, n_expected_observed, n_missing, pct_missing) %>%
  arrange(desc(n_expected_observed), pct_missing)

# these are the most 'complete' sites, with the longest observations and the fewest missing
head(missing_by_site, 10)
```






```{r impute missing}
# For each site (row), linearly interpolate only between the first and last
# observed (non-NA) value. NAs outside that window remain NA (site offline).
impute_row <- function(row) {
  obs <- which(!is.na(row))
  if (length(obs) < 2L) return(row)
  first <- obs[1L]; last <- obs[length(obs)]
  row[first:last] <- approx(obs, row[obs], xout = first:last)$y
  row
}
```



# Change-point detection for WBE

Defining a meaningful state of control is the central challenge of WBE.
These data streams are subject to well-documented sources of noise and systematic
uncertainty [Joung et al (2023)](https://pmc.ncbi.nlm.nih.gov/articles/PMC10890812/pdf/CCDR-49-166.pdf):

- **Target quantity**: The true target quantity of interest is the average viral load in a particular population week over week. This quantity varies naturally over time when the virus is in an endemic state. While this variation is not genuinely random, it can be modeled as such. For instance, we might assume it maintains some low-level average with as-if-Gaussian fluctuations from week to week, in which case the sequence is _stationary_ in the endemic phase.
- **Sampling**: The methods of sample collection, shipping, and storage (compositing vs grab sampling, location, timing, shipping temperature, distance to lab, missing data), which may vary across sites and time. The sites that were selected for our purposes primarily used composite sampling.
- **Analysis**: Laboratory analysis variability (de-gritting subsampling, homogenization, instrument error, human error).
- **Infrastructure**: The location and engineering of sewage systems and wastewater treatment plants (catchment area size and population, combined or separate ('sanitary') collection of runoff and wastewater).
- **Weather**: Weather conditions (rainwater or snowmelt contributing to influent, temperature, pH), which vary across sites and seasons, with some degree of stochasticity.

In the literature, these uncertainties are dealt with under the rubric of _normalization_, which is an active area of research.
In some sense, defining a proper state of control from which deviations are meaningful is the whole challenge of WBE.
It may be easier to define the state of control for certain sites, like those with sanitary sewage systems that are less influenced by major weather events.



---

# Detection pipeline: localized alarm on normalized de-seasonalized signal

This section implements the pipeline from the slides: normalize (COVID/RSV ratio),
de-seasonalize (cyclic spline on 2023–2025 training data), and run a
`LocalizedDetector` on the probit-transformed residuals from the 2025–2026
held-out period.

## Site selection

We redefine `selected_sites` with explicit inclusion criteria: enrolled by
2023-01-01, observed through 2026-01-01, and ≥ 80% complete data for both
covN2 and RSV within that window.

```{r pipeline-site-selection}
library(mgcv)

window_start <- as.Date("2023-01-01")
train_end    <- as.Date("2025-01-01")
window_end   <- as.Date("2026-01-01")
spine_pipe   <- seq(window_start, window_end, by = "7 days")
n_expected   <- length(spine_pipe)

# Enrollment window from full raw data (covN2 used as the enrollment marker)
site_enrollment <- ww %>%
  filter(measureid == "covN2", site != "") %>%
  group_by(site) %>%
  summarize(first_obs = min(weekstart), last_obs = max(weekstart), .groups = "drop")

# Completeness for both covN2 and RSV within the evaluation window
site_completeness <- ww %>%
  filter(measureid %in% c("covN2", "rsv"), site != "",
         weekstart >= window_start, weekstart <= window_end) %>%
  select(site, weekstart, measureid, w_avg) %>%
  pivot_wider(names_from = measureid, values_from = w_avg) %>%
  group_by(site) %>%
  summarize(
    n_covN2   = sum(!is.na(covN2)),
    n_rsv     = sum(!is.na(rsv)),
    pct_covN2 = round(n_covN2 / n_expected, 3),
    pct_rsv   = round(n_rsv   / n_expected, 3),
    .groups   = "drop"
  ) %>%
  left_join(site_enrollment, by = "site") %>%
  mutate(
    enrolled_by_start = first_obs <= window_start,
    through_end       = last_obs  >= window_end,
    meets_criteria    = enrolled_by_start & through_end &
                        pct_covN2 >= 0.80 & pct_rsv >= 0.80
  )

# Report sites that fail
cat("=== Sites failing inclusion criteria ===\n")
site_completeness %>%
  filter(!meets_criteria) %>%
  arrange(pct_rsv) %>%
  select(site, first_obs, last_obs, pct_covN2, pct_rsv,
         enrolled_by_start, through_end) %>%
  knitr::kable()

# Redefine selected_sites
selected_sites <- site_completeness %>%
  filter(meets_criteria) %>%
  arrange(site) %>%
  pull(site)

cat("\nSelected sites (n =", length(selected_sites), "):\n")
cat(paste(selected_sites, collapse = "\n"), "\n")
```

## Data preparation: log(covN2 / RSV), imputation

```{r pipeline-data-prep}
K_sites <- length(selected_sites)

pipe_long <- ww %>%
  filter(measureid %in% c("covN2", "rsv"), site %in% selected_sites) %>%
  select(site, weekstart, measureid, w_avg) %>%
  pivot_wider(names_from = measureid, values_from = w_avg) %>%
  filter(weekstart >= window_start, weekstart <= window_end) %>%
  # Drop low covN2 outliers (no lower bound applied to RSV)
  mutate(covN2 = ifelse(!is.na(covN2) & covN2 < 0.01, NA, covN2)) %>%
  mutate(log_ratio = ifelse(covN2 > 0 & !is.na(covN2) & rsv > 0 & !is.na(rsv),
                             log(covN2 / rsv), NA_real_)) %>%
  select(site, weekstart, log_ratio) %>%
  # Fill the full weekly spine and interpolate gaps within each site's observed window
  complete(site, weekstart = spine_pipe) %>%
  arrange(site, weekstart) %>%
  group_by(site) %>%
  mutate(log_ratio = impute_row(log_ratio)) %>%
  ungroup()

cat("NAs remaining after imputation:", sum(is.na(pipe_long$log_ratio)),
    "(leading/trailing offline periods)\n")

# Wide matrix: K_sites rows x n_weeks columns
ratio_matrix <- pipe_long %>%
  pivot_wider(names_from = weekstart, values_from = log_ratio) %>%
  column_to_rownames("site") %>%
  as.matrix()
```

## De-seasonalization: cyclic spline on 2023–2025

```{r pipeline-deseasonalize}
train_idx <- which(spine_pipe <  train_end)
test_idx  <- which(spine_pipe >= train_end)

# Fit cyclic cubic spline on day-of-year using 2023–2025 training data;
# predict the seasonal curve over the full window (2023–2026) to remove it.
deseason_site <- function(y, dates) {
  df <- data.frame(y = y, date = dates,
                   yday = as.integer(format(dates, "%j")))
  fit <- mgcv::gam(y ~ s(yday, bs = "cc", k = 10),
                   data   = df[dates < train_end, ],
                   knots  = list(yday = c(1, 365)),
                   na.action = na.omit)
  y - predict(fit, newdata = df)
}

resid_matrix <- ratio_matrix
for (k in seq_len(K_sites))
  resid_matrix[k, ] <- deseason_site(ratio_matrix[k, ], spine_pipe)

# Quick sanity plot: training residuals should be roughly mean-zero
train_resid_df <- as.data.frame(resid_matrix[, train_idx]) %>%
  rownames_to_column("site") %>%
  pivot_longer(-site, names_to = "date_chr", values_to = "resid") %>%
  mutate(date = as.Date(date_chr))

ggplot(train_resid_df, aes(x = date, y = resid, colour = site)) +
  geom_line(linewidth = 0.4, alpha = 0.7, na.rm = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ site, ncol = 3, scales = "free_y") +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b '%y") +
  labs(x = NULL, y = "Residual log(covN2/RSV)",
       title = "Training-period de-seasonalized residuals (2023–2025)") +
  theme_bw(base_size = 9) +
  theme(axis.text.x      = element_text(angle = 30, hjust = 1),
        strip.background = element_rect(fill = "#D6EAF8"),
        strip.text       = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())
```

## Probit transform and BoundedModel increments

```{r pipeline-probit}
# Standardize residuals by training-period SD per site; apply probit transform.
# Under H0 (residuals ~ N(0, sigma^2)), pnorm(resid/sd) ~ Uniform(0,1) with
# conditional mean 0.5, satisfying the BoundedModel null E[X_t] <= eta = 0.5.
train_sd      <- apply(resid_matrix[, train_idx, drop = FALSE], 1, sd, na.rm = TRUE)
probit_matrix <- pnorm(sweep(resid_matrix, 1, train_sd, "/"))

# Test-period matrix: N_test x K_sites (as required by LocalizedDetector)
test_mat   <- t(probit_matrix[, test_idx, drop = FALSE])
test_dates <- spine_pipe[test_idx]
N_test     <- nrow(test_mat)

# Kelly-optimal fixed bet for a Beta(3,1) post-change distribution with eta = 0.5.
# Discretise the Beta(3,1) density on a fine grid and pass to
# compute_kelly_optimal_bet(), which finds lambda* maximising
# E_{Beta(3,1)}[log(1 + lambda*(X - eta))].
x_grid           <- seq(0.01, 0.99, length.out = 2000)
kelly_bet_beta <- compute_kelly_optimal_bet(
  l_i = x_grid,
  eta = 0.5,
  p_i = dbeta(x_grid, 1.5, 1)
)

# BoundedModel log-increments per stream using the Kelly-optimal fixed bet.
inc_mat <- matrix(NA_real_, nrow = N_test, ncol = K_sites,
                  dimnames = list(NULL, rownames(probit_matrix)))
for (k in seq_len(K_sites))
  inc_mat[, k] <- compute_increments(
    TSM(BoundedModel(eta = 0.5, bets = FixedBet(kelly_bet_beta))),
    test_mat[, k], log = TRUE)
```

```{r beta31-kelly-plot}
# Beta(3,1) density annotated with the Kelly-optimal bet.
eta_val  <- 0.5
lam_star <- kelly_bet_beta

df_beta <- data.frame(x = x_grid, density = dbeta(x_grid, 1.5, 1))

fig_dir <- file.path(
  dirname(dirname(knitr::current_input(dir = TRUE))),
  "ssc-2026", "figures", "nwmp"
)

p_beta <- ggplot(df_beta, aes(x = x, y = density)) +
  geom_area(data = subset(df_beta, x >= eta_val),
            fill = "#2E86AB", alpha = 0.25) +
  geom_area(data = subset(df_beta, x < eta_val),
            fill = "#E84855", alpha = 0.25) +
  geom_line(linewidth = 0.9, colour = "#1a1a2e") +
  geom_vline(xintercept = eta_val, linetype = "dashed",
             colour = "grey40", linewidth = 0.6) +
  annotate("text", x = eta_val + 0.02, y = 2.8,
           label = "eta == 0.5", parse = TRUE, hjust = 0, size = 3.5) +
  annotate("text", x = 0.76, y = 1.0,
           label = paste0("Kelly-optimal bet\nλ* = ", round(lam_star, 4)),
           hjust = 0, size = 3.2, colour = "#2E86AB") +
  scale_x_continuous(breaks = seq(0, 1, 0.25)) +
  labs(x = "x", y = "Density",
       title = "Beta(3/2, 1) post-change distribution",
       subtitle = paste0("Kelly-optimal bet λ* = 1/η = ", round(lam_star, 4), "")) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())

print(p_beta)
ggsave(file.path(fig_dir, "beta_kelly_bet.png"), p_beta,
       width = 6, height = 4, dpi = 300)
```

## Raw vs. transformed streams

```{r pipeline-compare-plot, fig.width=12, fig.height=9}
library(patchwork)

resid_long <- as.data.frame(resid_matrix) %>%
  rownames_to_column("site") %>%
  pivot_longer(-site, names_to = "date_chr", values_to = "resid") %>%
  mutate(date = as.Date(date_chr))

probit_long <- as.data.frame(probit_matrix) %>%
  rownames_to_column("site") %>%
  pivot_longer(-site, names_to = "date_chr", values_to = "probit") %>%
  mutate(date = as.Date(date_chr))

p_raw <- resid_long %>%
  ggplot(aes(x = date, y = resid, colour = site)) +
  geom_line(linewidth = 0.4, alpha = 0.8, na.rm = TRUE) +
  geom_vline(xintercept = train_end, linetype = "dashed", colour = "grey50") +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey40") +
  facet_wrap(~ site, ncol = 2, scales = "free_y") +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = "De-seasonalized residual",
       title = "De-seasonalized log-ratio residuals",
       subtitle = "Dashed = train / test split (Jan 2025)") +
  theme_bw(base_size = 8) +
  theme(axis.text.x      = element_text(angle = 30, hjust = 1),
        strip.background = element_rect(fill = "#D6EAF8"),
        strip.text       = element_text(face = "bold", size = 7),
        panel.grid.minor = element_blank())

p_probit <- probit_long %>%
  ggplot(aes(x = date, y = probit, colour = site)) +
  geom_line(linewidth = 0.4, alpha = 0.8, na.rm = TRUE) +
  geom_vline(xintercept = train_end, linetype = "dashed", colour = "grey50") +
  geom_hline(yintercept = 0.5, linetype = "dotted", colour = "grey40") +
  facet_wrap(~ site, ncol = 2, scales = "free_y") +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = NULL, y = "Probit-transformed residual",
       title = "Probit(de-seasonalized residual)",
       subtitle = "Dotted = null mean (0.5)") +
  theme_bw(base_size = 8) +
  theme(axis.text.x      = element_text(angle = 30, hjust = 1),
        strip.background = element_rect(fill = "#FDEBD0"),
        strip.text       = element_text(face = "bold", size = 7),
        panel.grid.minor = element_blank())

p_raw | p_probit
```

## LocalizedDetector on 2025–2026

```{r pipeline-detect}
alpha_global <- 1/(52 * 5)
log_thresh   <- log(1 / alpha_global)

# Bonferroni allowance: uniform 1/K per stream
ld      <- LocalizedDetector(K = K_sites, alpha = alpha_global, criterion = "ARL")
det_out <- run_detector(ld, evidence = inc_mat, log = TRUE)

# Per-stream stopping times → calendar dates
stream_stops <- vapply(det_out$stream_results, `[[`, numeric(1L), "stopping_time")
stop_dates   <- ifelse(is.finite(stream_stops),
                        format(test_dates[pmin(as.integer(stream_stops), N_test)]),
                        "no alarm")
results_df <- data.frame(site      = rownames(probit_matrix),
                          stop_date = stop_dates,
                          row.names = NULL)

cat("=== Detection results (alpha =", alpha_global,
    ", threshold = log(1/alpha) =", round(log_thresh, 2), ") ===\n")
cat("Global alarm:", det_out$alarm, "\n")
if (det_out$alarm) {
  cat("Global stopping time:",
      format(test_dates[det_out$stopping_time]), "\n")
  cat("First alarming stream:",
      rownames(probit_matrix)[det_out$first_alarm_stream], "\n")
}
cat("\nPer-stream results:\n")
print(results_df)
```

## Alarm plot

```{r pipeline-plot, fig.height=9}
site_names <- rownames(probit_matrix)

# Alarm annotation data frame (streams that fired)
alarm_df <- results_df %>%
  filter(stop_date != "no alarm") %>%
  mutate(alarm_date = as.Date(stop_date))

# Log S-R statistic trajectories from the detector output
sr_df <- lapply(seq_len(K_sites), function(k) {
  data.frame(date   = test_dates,
             site   = site_names[k],
             log_SR = det_out$stream_results[[k]]$statistic)
}) %>% bind_rows()

# Probit residuals for test period (long format)
probit_test_df <- as.data.frame(t(test_mat)) %>%
  tibble::rownames_to_column("site") %>%
  tidyr::pivot_longer(-site, names_to = "date_chr", values_to = "probit") %>%
  dplyr::mutate(date = as.Date(date_chr))

# Panel 1: probit residuals with null-mean reference and alarm lines
p_raw <- ggplot(probit_test_df, aes(x = date, y = probit, colour = site)) +
  geom_line(linewidth = 0.5, alpha = 0.85, na.rm = TRUE) +
  geom_hline(yintercept = 0.5, linetype = "dotted", colour = "grey40") +
  geom_vline(data = alarm_df, aes(xintercept = alarm_date, colour = site),
             linetype = "dashed", linewidth = 0.75) +
  facet_wrap(~ site, ncol = 3) +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(x = NULL, y = "Probit residual",
       title = "Probit-transformed residuals — 2025–2026 test period",
       subtitle = "Dotted = null mean (0.5); dashed = first S-R alarm") +
  theme_bw(base_size = 9) +
  theme(strip.background = element_rect(fill = "#D6EAF8"),
        strip.text       = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())

# Panel 2: log S-R statistic with threshold
p_sr <- ggplot(sr_df, aes(x = date, y = log_SR, colour = site)) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  geom_hline(yintercept = log_thresh, linetype = "dashed",
             colour = "black", linewidth = 0.7) +
  geom_vline(data = alarm_df, aes(xintercept = alarm_date, colour = site),
             linetype = "dashed", linewidth = 0.75) +
  facet_wrap(~ site, ncol = 3) +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(x = NULL, y = expression(log~R[t]),
       title = sprintf("Log Shiryaev–Roberts statistic (α = %.2f, threshold = %.1f)",
                       alpha_global, log_thresh)) +
  theme_bw(base_size = 9) +
  theme(strip.background = element_rect(fill = "#FDEBD0"),
        strip.text       = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())

print(p_raw)
print(p_sr)
```

---

## Simple threshold baseline

Before the statistical detector, it is worth asking what a completely
transparent, rule-based approach would find: raise an alarm at site $k$
the first time its log-covN2 signal exceeds the 95th percentile of its
own 2023–2025 baseline.  The two chunks below first display the training
period so the baseline distribution is visible, then apply the rule to
the 2025–2026 hold-out period.

```{r threshold-baseline-train, fig.height=8, fig.cap="Log-covN2 signal at selected sites during the 2023–2025 training period. The red dashed line marks the site-specific 95th percentile used as the detection threshold."}
# Raw log-covN2 (already log10-transformed at source) for selected sites
train_cov <- ww %>%
  filter(measureid == "covN2", site %in% selected_sites,
         weekstart >= window_start, weekstart < train_end) %>%
  select(site, weekstart, w_avg)

# Site-specific 95th-percentile threshold from training data
site_p95 <- train_cov %>%
  group_by(site) %>%
  summarize(p95 = quantile(w_avg, 0.95, na.rm = TRUE), .groups = "drop")

ggplot(train_cov, aes(x = weekstart, y = w_avg, colour = site)) +
  geom_line(linewidth = 0.5, alpha = 0.85, na.rm = TRUE) +
  geom_hline(data = site_p95, aes(yintercept = p95),
             linetype = "dashed", colour = "firebrick", linewidth = 0.7) +
  facet_wrap(~ site, ncol = 3, scales = "free_y") +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b\n'%y") +
  labs(x = NULL, y = expression(log[10]~"(copies / mL)"),
       title = "Training period: log-covN2 at selected sites (2023–2025)",
       subtitle = "Red dashed line = site-specific 95th percentile (detection threshold)") +
  theme_bw(base_size = 9) +
  theme(strip.background = element_rect(fill = "#D6EAF8"),
        strip.text       = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())
```

```{r threshold-baseline-test, fig.height=8, fig.cap="Log-covN2 signal during the 2025–2026 test period. The red dashed line is the 95th-percentile threshold from 2023–2025 training data; the vertical orange line marks the first week each site exceeds the threshold."}
test_cov <- ww %>%
  filter(measureid == "covN2", site %in% selected_sites,
         weekstart >= train_end, weekstart <= window_end) %>%
  select(site, weekstart, w_avg)

# First week each site exceeds the training-period 95th percentile
first_exceed <- test_cov %>%
  left_join(site_p95, by = "site") %>%
  filter(!is.na(w_avg), w_avg > p95) %>%
  group_by(site) %>%
  slice_min(weekstart, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(alarm_date = weekstart)

cat("=== Simple threshold detections (95th percentile rule) ===\n")
all_sites_df <- data.frame(site = selected_sites) %>%
  left_join(first_exceed %>% select(site, alarm_date), by = "site") %>%
  mutate(alarm_date = ifelse(is.na(alarm_date), "no alarm",
                             format(alarm_date)))
print(all_sites_df)

ggplot(test_cov, aes(x = weekstart, y = w_avg, colour = site)) +
  geom_line(linewidth = 0.5, alpha = 0.85, na.rm = TRUE) +
  geom_hline(data = site_p95, aes(yintercept = p95),
             linetype = "dashed", colour = "firebrick", linewidth = 0.7) +
  geom_vline(data = first_exceed, aes(xintercept = alarm_date),
             colour = "darkorange", linewidth = 0.9) +
  facet_wrap(~ site, ncol = 3, scales = "free_y") +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(x = NULL, y = expression(log[10]~"(copies / mL)"),
       title = "Test period: log-covN2 with 95th-percentile threshold alarms (2025–2026)",
       subtitle = "Red dashed = training-period 95th percentile; orange = first exceedance") +
  theme_bw(base_size = 9) +
  theme(strip.background = element_rect(fill = "#D6EAF8"),
        strip.text       = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())
```

---

# Secondary analysis: raw CovN2 with AGRAPA betting

The full pipeline above involves several normalization steps (RSV ratio,
seasonal adjustment, probit transform) whose purpose is to manufacture a
stationary, mean-zero null against which the BoundedModel TSM is valid.
Here we ask what happens if we skip all of that and feed the raw
log-covN2 signal directly into the same betting machinery.

The pre-change null mean $\eta_k$ for each site is estimated as the
**90th percentile** of the site's normalized training-period values.
Setting $\eta$ to a high quantile is deliberately conservative: it
places the null boundary at a level that the data routinely reached
before 2025, so the martingale only grows when the signal persistently
**exceeds** the historical upper tail.  The EWMA-AGRAPA bet adapts its
aggressiveness based on recent data, with a soft window of roughly 10
weeks ($\rho = 0.1$).

```{r raw-covN2-prep}
# Extract raw log10-covN2 for selected sites over the full window,
# filling the weekly spine and linearly interpolating interior gaps.
raw_long <- ww %>%
  filter(measureid == "covN2", site %in% selected_sites,
         weekstart >= window_start, weekstart <= window_end) %>%
  select(site, weekstart, w_avg) %>%
  complete(site, weekstart = spine_pipe) %>%
  arrange(site, weekstart) %>%
  group_by(site) %>%
  mutate(w_avg = impute_row(w_avg)) %>%
  ungroup()

# Wide matrix: K_sites rows x n_weeks columns
raw_mat <- raw_long %>%
  pivot_wider(names_from = weekstart, values_from = w_avg) %>%
  column_to_rownames("site") %>%
  as.matrix()

# Normalize each site by its training-period maximum so that training
# values are in [0, 1]; test values exceeding the max are clamped to 1.
train_max_raw <- apply(raw_mat[, train_idx, drop = FALSE], 1, max, na.rm = TRUE)
norm_raw_mat  <- pmin(sweep(raw_mat, 1, train_max_raw, "/"), 1)

# Per-site eta: 90th percentile of normalized training values.
# Under H0 (pre-change), E[X_t] << 90th pctile for right-skewed CovN2,
# so the BoundedModel null E[X_t] <= eta is well satisfied.
eta_raw <- apply(norm_raw_mat[, train_idx, drop = FALSE], 1,
                 quantile, probs = 0.90, na.rm = TRUE)

cat("Per-site eta (90th pctile of normalized training CovN2):\n")
print(round(eta_raw, 3))
```

```{r raw-covN2-detect}
test_raw_mat <- t(norm_raw_mat[, test_idx, drop = FALSE])   # N_test x K
N_raw        <- nrow(test_raw_mat)

# BoundedModel log-increments per stream using site-specific eta
# and EWMA-AGRAPA betting (rho = 0.1 ~ 10-week soft window).
inc_raw <- matrix(NA_real_, nrow = N_raw, ncol = K_sites,
                  dimnames = list(NULL, selected_sites))
for (k in seq_len(K_sites))
  inc_raw[, k] <- compute_increments(
    TSM(BoundedModel(eta = eta_raw[k], bets = EWMABet(rho = 0.1))),
    test_raw_mat[, k], log = TRUE)

# LocalizedDetector with the same alpha and ARL criterion as the main pipeline.
ld_raw  <- LocalizedDetector(K = K_sites, alpha = alpha_global, criterion = "ARL")
det_raw <- run_detector(ld_raw, evidence = inc_raw, log = TRUE)

stream_stops_raw <- vapply(det_raw$stream_results, `[[`, numeric(1L), "stopping_time")
stop_dates_raw   <- ifelse(
  is.finite(stream_stops_raw),
  format(test_dates[pmin(as.integer(stream_stops_raw), N_raw)]),
  "no alarm"
)
results_raw_df <- data.frame(site      = selected_sites,
                             stop_date = stop_dates_raw,
                             row.names = NULL)

cat("=== Raw CovN2 detection (EWMA-AGRAPA, alpha =", alpha_global,
    ", threshold =", round(log_thresh, 2), ") ===\n")
cat("Global alarm:", det_raw$alarm, "\n")
if (det_raw$alarm)
  cat("Global stopping time:", format(test_dates[det_raw$stopping_time]), "\n")
cat("\nPer-stream results:\n")
print(results_raw_df)
```

```{r raw-covN2-plot, fig.height=9}
alarm_raw_df <- results_raw_df %>%
  filter(stop_date != "no alarm") %>%
  mutate(alarm_date = as.Date(stop_date))

# Log S-R trajectories
sr_raw_df <- lapply(seq_len(K_sites), function(k) {
  data.frame(date   = test_dates,
             site   = selected_sites[k],
             log_SR = det_raw$stream_results[[k]]$statistic)
}) %>% bind_rows()

# Normalized CovN2 in test period (long format), with per-site eta for reference
eta_df <- data.frame(site = selected_sites, eta = eta_raw)

norm_test_df <- as.data.frame(t(norm_raw_mat[, test_idx, drop = FALSE])) %>%
  tibble::rownames_to_column("date_chr") %>%
  tidyr::pivot_longer(-date_chr, names_to = "site", values_to = "value") %>%
  dplyr::mutate(date = as.Date(date_chr)) %>%
  left_join(eta_df, by = "site")

# Panel 1: normalized CovN2 with per-site eta and alarm lines
p_raw2 <- ggplot(norm_test_df, aes(x = date, y = value, colour = site)) +
  geom_line(linewidth = 0.5, alpha = 0.85, na.rm = TRUE) +
  geom_hline(data = eta_df, aes(yintercept = eta),
             linetype = "dotted", colour = "grey30", linewidth = 0.6) +
  geom_vline(data = alarm_raw_df, aes(xintercept = alarm_date, colour = site),
             linetype = "dashed", linewidth = 0.75) +
  facet_wrap(~ site, ncol = 3) +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(x = NULL, y = "Normalized log-covN2",
       title = "Raw CovN2 (normalized to training max) — 2025–2026 test period",
       subtitle = "Dotted = per-site eta (90th pctile training); dashed = first S-R alarm") +
  theme_bw(base_size = 9) +
  theme(strip.background = element_rect(fill = "#D6EAF8"),
        strip.text       = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())

# Panel 2: log S-R statistic with threshold
p_sr2 <- ggplot(sr_raw_df, aes(x = date, y = log_SR, colour = site)) +
  geom_line(linewidth = 0.5, alpha = 0.85) +
  geom_hline(yintercept = log_thresh, linetype = "dashed",
             colour = "black", linewidth = 0.7) +
  geom_vline(data = alarm_raw_df, aes(xintercept = alarm_date, colour = site),
             linetype = "dashed", linewidth = 0.75) +
  facet_wrap(~ site, ncol = 3) +
  scale_colour_viridis_d(option = "turbo", guide = "none") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(x = NULL, y = expression(log~R[t]),
       title = sprintf("Log S-R statistic — raw CovN2 AGRAPA (alpha = %.4f, threshold = %.1f)",
                       alpha_global, log_thresh)) +
  theme_bw(base_size = 9) +
  theme(strip.background = element_rect(fill = "#FDEBD0"),
        strip.text       = element_text(face = "bold", size = 8),
        panel.grid.minor = element_blank())

print(p_raw2)
print(p_sr2)
```
