---
title: "multichangepoints workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multichangepoints workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

This is a quick-start guide to the three core building blocks of `multichangepoints`:
**global testing** (is anything changing, anywhere?), **localization** (which streams are
changing?), and **graph-aware localization** (using a proximity structure between streams
to localize faster). All three examples use the same kind of data -- independent Gaussian
streams -- and each is a self-contained pair of chunks: one generates data, the next
applies the package's tools to it.

```{r setup}
devtools::load_all("~/Dropbox/CANSSI/multi-change-points/")
library(multichangepoints)
set.seed(20260305)
```

## 1. Global testing via product combining

Product combining multiplies K independent marginal test supermartingales (TSMs) into one
joint TSM, and is the most efficient way to combine streams when they really are
independent (`ProductCombiner`). We simulate K = 6 IID Gaussian streams that all shift at
the same change-point, and ask a single global question: has *anything* changed?

```{r global-data}
set.seed(1)
K  <- 6
N  <- 150
nu <- 50   # change-point

# all K streams shift mean 0 -> 1 at nu
x_global <- matrix(rnorm(N * K), nrow = N, ncol = K)
x_global[(nu + 1):N, ] <- matrix(rnorm((N - nu) * K, mean = 1), nrow = N - nu)
```

```{r global-detect}
# one marginal TSM per stream (true pre/post means, shared unit variance)
marg_tsm <- TSM(GaussianModel(mean_pre = 0, sd_pre = 1, mean_post = 1, sd_post = 1))

marg_inc <- sapply(seq_len(K), function(k)
  compute_increments(marg_tsm, x_global[, k], log = TRUE))

# combine into one joint TSM, then run an ARL-controlled S-R detector on it
inc_joint <- combine_streams(ProductCombiner(), marg_inc, log = TRUE)
sr        <- ShiryaevRobertsDetector(alpha = 0.01, criterion = "ARL")
det_joint <- run_detector(sr, inc_joint, log = TRUE)

cat("Global alarm:", is.finite(det_joint$stopping_time), "\n")
cat("Stopping time:", det_joint$stopping_time,
    " (delay", det_joint$stopping_time - nu, "steps)\n")
```

## 2. Localization with a uniform spending allowance

`LocalizedDetector` runs one marginal detector per stream and splits a shared error budget
across them; with no further arguments the budget is split uniformly (1/K each step), which
is exactly the e-d-Bonferroni procedure. Because each stream is watched on its own, this
also identifies *which* stream changed -- something the combined statistic above cannot do,
and something product combining is not even well suited to detect when, as here, only one
of the K streams actually changes (the other K-1 null streams' log-increments pull the
combined statistic down on average).

```{r loc-uniform-data}
set.seed(1)
# only stream 1 shifts at nu; streams 2-6 stay N(0, 1) throughout
x_loc <- matrix(rnorm(N * K), nrow = N, ncol = K)
x_loc[(nu + 1):N, 1] <- rnorm(N - nu, mean = 1)
```

```{r loc-uniform-detect}
marg_inc_loc <- sapply(seq_len(K), function(k)
  compute_increments(marg_tsm, x_loc[, k], log = TRUE))

ld_unif  <- LocalizedDetector(K = K, alpha = 0.01, criterion = "ARL")  # allowance = 1/K
det_unif <- run_detector(ld_unif, marg_inc_loc, log = TRUE)

stop_unif <- sapply(det_unif$stream_results, function(r) r$stopping_time)
cat("Per-stream stopping times:\n")
print(stop_unif)
cat("Alarmed streams:", which(is.finite(stop_unif)), "\n")
```

## 3. Graph-aware localization

When streams sit on a proximity graph, a **graph-aware allowance** redirects the budget
toward the not-yet-alarmed neighbors of streams that have already fired, rather than
spreading it uniformly. We build a small 6-node chain (`ProximityGraph`), let a change
propagate outward from one end (`propagation_nu`), and simulate each stream shifting at its
*own* change-point.

```{r loc-graph-data}
set.seed(4)
# a 6-node chain: 1 -- 2 -- 3 -- 4 -- 5 -- 6, unit edge weights
W <- matrix(0, K, K)
for (k in seq_len(K - 1)) W[k, k + 1] <- W[k + 1, k] <- 1
prox_graph <- ProximityGraph(W)

# change starts at node 1 (nu0 = 50) and propagates one step every 20 time units,
# so nu_k = 50, 70, 90, 110, 130, 150 for k = 1, ..., 6
nu_vec <- propagation_nu(prox_graph, source = 1, nu0 = 50, scale = 20)

N_g <- 220
x_graph <- matrix(rnorm(N_g * K), nrow = N_g, ncol = K)
for (k in seq_len(K)) {
  post_idx <- (nu_vec[k] + 1):N_g
  x_graph[post_idx, k] <- rnorm(length(post_idx), mean = 1)
}
```

```{r loc-graph-detect}
marg_inc_graph <- sapply(seq_len(K), function(k)
  compute_increments(marg_tsm, x_graph[, k], log = TRUE))

ld_graph <- LocalizedDetector(
  K = K, alpha = 0.01, criterion = "ARL",
  proximity_graph = prox_graph,
  kernel          = exponential_kernel(xi = 1.5),
  zeta            = 0.8
)
det_graph <- run_detector(ld_graph, marg_inc_graph, log = TRUE)

stop_graph <- sapply(det_graph$stream_results, function(r) r$stopping_time)
cat("True change-points:      ", nu_vec, "\n")
cat("Per-stream stopping times:", stop_graph, "\n")
```

Because the allowance keeps shifting toward the neighbors of already-alarmed nodes, each
stream tends to be caught at least as fast -- often a little faster -- as it would be under
the uniform allowance of Section 2, without ever compromising the global ARL guarantee. On
a single small chain like this one the improvement per stream is modest; run at scale,
across larger and more structured graphs, the same comparison is the subject of the
package's localization simulation study, where the gap becomes substantial.

## Where to go from here

This guide covers the minimal path through the package, but every piece of it is
swappable:

- **Other models.** Any marginal TSM works in place of `GaussianModel` above --
  `BoundedModel` gives a distribution-free, AGRAPA-betting TSM for approximately bounded
  data (no need to specify a post-change distribution); `GaussianVARModel` and
  `ARBoundedModel` handle autocorrelated streams; `MultivariateGaussianModel` and
  `BernoulliModel` cover other data types.
- **Dependency graphs and other ways to combine.** `ProductCombiner` (Section 1) assumes
  full independence. `AverageCombiner` and `UniversalPortfolioCombiner` stay valid under
  arbitrary dependence, at some cost in efficiency. When only *some* streams are dependent,
  a `DependencyGraph` plus `GraphCombiner` averages within each dependent block and
  multiplies across blocks, recovering the best of both.
- **Other graph-aware allowances.** `exponential_kernel(xi)` (Section 3) is one proximity
  kernel; `hard_cutoff_kernel(r)` is another. `zeta` moves continuously between the uniform
  allowance (`zeta = 0`) and one that concentrates almost entirely on the active
  neighborhood (`zeta` near 1). Any other topology -- hub-and-spoke, disconnected clusters,
  a graph built from real distances or transit times -- is just a different weight matrix
  passed to `ProximityGraph`. A fixed, non-adaptive `allowance` vector can also be supplied
  directly to `LocalizedDetector` in place of a graph, e.g. to encode prior knowledge about
  which streams are more likely to change.
- **PFA control.** Every detector above targets the average run length
  (`criterion = "ARL"`). Passing `criterion = "PFA"` together with a `spending` schedule
  (a vector summing to 1, e.g. from a geometric distribution, or a full time-by-stream
  matrix) instead caps the *probability* of ever falsely alarming, at the cost of requiring
  the spending schedule to concentrate its budget near the true change-point for good power.
