---
title: "multichangepoints workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multichangepoints workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

This vignette demonstrates the core workflow for sequential change-point detection with:

- IID Gaussian data generating process (DGP)
- simple-vs-simple test supermartingale (TSM)
- Shiryaev-Roberts (S-R) detector

```{r setup}
devtools::load_all("~/Dropbox/CANSSI/multi-change-points/")
library(multichangepoints)
set.seed(20260305)
```

## 1. Specify a Gaussian model and TSM

```{r}
# define a simple change model, including all pre-change and post-change parameters
model <- GaussianModel(
  mean_pre = 0,
  sd_pre = 1,
  mean_post = 1,
  sd_post = 1
)

# construct a tsm object from the model, it will receive data
tsm <- SimpleVsSimpleTSM(model)
```


## 2. Create an IID Gaussian DGP

```{r}
# set up the data generating process for the data sequence, including the pre/post distributions and the change-point
dgp <- DGP(
  generator = default_gaussian_dgp,
  pre_params = list(mean = 0, sd = 1), # parameters pre-change
  post_params = list(mean = 1, sd = 1), # parameters post-change
  nu = 120, # change-point
  name = "iid-gaussian-change"
)

# generate data from the dgp
N <- 250
x <- generate_stream(dgp, N = N, K = 1)
head(x)
```


## 3. Build increments and run the S-R detector with ARL control

```{r}
# increment sequence from data x
inc <- compute_increments(tsm, x)

# instantiate a detector class with ARL control
sr_1 <- ShiryaevRobertsDetector(
  alpha = 0.05, # average run length without false alarm is only 20
  criterion = "ARL"
)
sr_2 <- ShiryaevRobertsDetector(
  alpha = 0.001, # average run length without false alarm is 1000
  criterion = "ARL"
)


# push increments through detector sr
det_out_1 <- run_detector(sr_1, inc)
det_out_2 <- run_detector(sr_2, inc)
str(det_out_1)
```

The detector returns:

- `statistic`: the S-R path over time
- `stopping_time`: first alarm time (or `Inf`)
- `alarm`: whether an alarm occurred by the horizon

The following figure plots the path of the TSM and ALR-controlled S-R statistic, along with the alarm time.

```{r fig.cap="TSM helper path and ARL-controlled S-R statistic with alarm time."}
par(mfrow = c(1, 3))

# helper: construct TSM path from increments for visualization
z <- increments_to_tsm(inc, log = TRUE)
plot(z, type = "l", lwd = 2, col = "steelblue",
     main = "log-TSM sequence", xlab = "time", ylab = "TSM")
abline(v = dgp@nu, lty = 2, col = "gray50")

plot(log(det_out_1$statistic), type = "l", lwd = 2, col = "firebrick",
     main = "log-S-R statistic (ARL = 20)", xlab = "time", ylab = "R_t")
abline(h = log(sr_1@threshold), lty = 3, col = "gray50")
abline(v = dgp@nu, lty = 2, col = "gray50")
if (is.finite(det_out_1$stopping_time)) {
  abline(v = det_out_1$stopping_time, lwd = 2, col = "darkgreen")
}

plot(log(det_out_2$statistic), type = "l", lwd = 2, col = "firebrick",
     main = "log-S-R statistic (ARL = 1000)", xlab = "time", ylab = "R_t")
abline(h = log(sr_2@threshold), lty = 3, col = "gray50")
abline(v = dgp@nu, lty = 2, col = "gray50")
if (is.finite(det_out_2$stopping_time)) {
  abline(v = det_out_2$stopping_time, lwd = 2, col = "darkgreen")
}

par(mfrow = c(1,1))
```



## 4. Build increments and run the S-R detector with PFA control

```{r}
# define a spending schedule
spend_1 <- dgeom(1:N, p = 0.1) # geometric spending schedule with p = 0.1
spend_2 <- dpois(1:N, lambda = 120) # poisson spending schedule centered at true nu
spend_1 <- spend_1/sum(spend_1) # normalize
spend_2 <- spend_2/sum(spend_2)  # normalize


# instantiate a detectors with PFA control
sr_pfa_1 <- ShiryaevRobertsDetector(
  alpha = 0.05,
  criterion = "PFA",
  spending = spend_1
)
sr_pfa_2 <- ShiryaevRobertsDetector(
  alpha = 0.05,
  criterion = "PFA",
  spending = spend_2
)


# push increments through detector sr
det_out_1 <- run_detector(sr_pfa_1, inc)
det_out_2 <- run_detector(sr_pfa_2, inc)
```


```{r fig.cap="TSM helper path and PFA-controlled S-R statistic with alarm time."}
old_par <- par(mfrow = c(1, 2))


plot(log(det_out_1$statistic), type = "l", lwd = 2, col = "firebrick",
     main = "log-S-R statistic", xlab = "time", ylab = "R_t")
abline(h = log(sr_pfa_1@threshold), lty = 3, col = "gray50")
abline(v = dgp@nu, lty = 2, col = "gray50")
if (is.finite(det_out_1$stopping_time)) {
  abline(v = det_out_1$stopping_time, lwd = 2, col = "darkgreen")
}


plot(log(det_out_2$statistic), type = "l", lwd = 2, col = "firebrick",
     main = "log-S-R statistic", xlab = "time", ylab = "R_t")
abline(h = log(sr_pfa_2@threshold), lty = 3, col = "gray50")
abline(v = dgp@nu, lty = 2, col = "gray50")
if (is.finite(det_out_2$stopping_time)) {
  abline(v = det_out_2$stopping_time, lwd = 2, col = "darkgreen")
}

par(old_par)
```

## 5. Estimate operating characteristics by simulation

```{r}
sr <- ShiryaevRobertsDetector(
  alpha = 0.001,
  criterion = "ARL"
)

mc <- run_simulation(
  detector = sr,
  tsm = tsm,
  dgp = dgp,
  n_rep = 100,
  N = 300,
  seed = 77
)

mc
```

This returns Monte Carlo summaries including:

- `false_alarm_prob`: proportion of runs alarming before (or at) the change
- `ARL`: average run length to alarm (censored at horizon)
- `ADD`: average detection delay after the change

## Multivariate detection

Simple-vs-simple multivariate detection. 
Note that when the alternative is mis-specified, the helper TSM does not grow and the detector can be very inefficient. 

```{r}
K <- 3 # number of streams
N <- 250 # length of streams
dgp_mvn <- DGP(
  generator = default_multivariate_gaussian_dgp,
  pre_params = list(mu = c(-1, 0, 1), Sigma = diag(K)), # parameters pre-change
  post_params = list(mu = c(-1, 1, 1), Sigma = diag(K)), # parameters post-change
  nu = 120, # change-point
  name = "mv-gauss-change"
)

# generate data from the dgp
x <- generate_stream(dgp_mvn, N = N, K = K)

# true alternative
model_true <- MultivariateGaussianModel(
  mu_pre = c(-1, 0, 1),
  Sigma_pre = diag(K),
  mu_post = c(-1, 1, 1),
  Sigma_post = diag(K)
)
# mis-specified alternative
model_miss <- MultivariateGaussianModel(
  mu_pre = c(-1, 0, 1),
  Sigma_pre = diag(K),
  mu_post = c(1, 1, 1),
  Sigma_post = diag(K)
)

# instantiate TSMs and compute increments
inc_true <- compute_increments(SimpleVsSimpleTSM(model_true), x, log = TRUE)
inc_miss <- compute_increments(SimpleVsSimpleTSM(model_miss), x, log = TRUE)

# instantiate detector and compute sequence
sr <-  ShiryaevRobertsDetector(alpha = 0.001, criterion = "ARL")
det_out_true <- run_detector(sr, inc_true, log = TRUE)
det_out_miss <- run_detector(sr, inc_miss, log = TRUE)
```

```{r fig.cap="Multivariate data, helper TSMs, and detector statistics (true vs mis-specified alternative)."}
old_par <- par(mfrow = c(1, 3))

# Panel 1: K data streams
matplot(x, type = "l", lty = 1, lwd = 1.5,
        xlab = "time", ylab = "x_tk", main = "Data Streams", ylim = c(-5, 5))
abline(v = dgp_mvn@nu, lty = 2, col = "gray40")
legend("topleft",
       legend = paste0("stream ", seq_len(ncol(x))),
       col = seq_len(ncol(x)), lty = 1, lwd = 1.5, cex = 0.8, bty = "n")

# Panel 2: helper TSMs from increments
tsm_true <- compute_tsm(SimpleVsSimpleTSM(model_true), x, log = TRUE)
tsm_miss <- compute_tsm(SimpleVsSimpleTSM(model_miss), x, log = TRUE)
plot(tsm_true, type = "l", lwd = 2, col = "darkgreen",
     xlab = "time", ylab = "log helper TSM", main = "Helper TSMs", ylim = c(-500, 0))
lines(tsm_miss, lwd = 2, col = "darkorange")
abline(v = dgp_mvn@nu, lty = 2, col = "gray40")
legend("topleft",
       legend = c("true model", "mis-specified model"),
       col = c("darkgreen", "darkorange"),
       lty = 1, lwd = 2, cex = 0.8, bty = "n")

# Panel 3: detector statistics
plot(det_out_true$statistic, type = "l", lwd = 2, col = "darkgreen",
     xlab = "time", ylab = "log R_t", main = "S-R Statistics")
lines(det_out_miss$statistic, lwd = 2, col = "darkorange")
abline(h = log(sr@threshold), lty = 3, col = "gray40")
abline(v = dgp_mvn@nu, lty = 2, col = "gray40")
if (is.finite(det_out_true$stopping_time)) {
  abline(v = det_out_true$stopping_time, lwd = 1.5, col = "darkgreen")
}
if (is.finite(det_out_miss$stopping_time)) {
  abline(v = det_out_miss$stopping_time, lwd = 1.5, col = "darkorange")
}
legend("topleft",
       legend = c("true model", "mis-specified model"),
       col = c("darkgreen", "darkorange"),
       lty = 1, lwd = 2, cex = 0.8, bty = "n")

par(old_par)
```


## Composite models

## Application to commodity prices

```{r}
library(quantmod)

getSymbols(c("PMAIZMTUSDM", "PSOYBUSDM", "PWHEAMTUSDM", "DCOILWTICO"), src = "FRED", 
           from = "2005-01-01", 
           to = "2026-01-01")

# monthly prices
prices <- apply.monthly(merge(PMAIZMTUSDM, PSOYBUSDM, PWHEAMTUSDM, DCOILWTICO), colMeans, na.rm = T)
colnames(prices) <- c("corn", "soybean", "wheat", "oil")
returns <- diff(log(prices))
plot(returns, legend.loc = "topright")
```



